Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-13 arxiv/intro.tex
finish intro

\begin{abstract}
	Hard parameter sharing is a widely used approach to learn jointly from multiple tasks in many applications, such as text classification. The idea is to use a shared feature space for all tasks, while each task also has a separate layer for making prediction.
%	It becomes increasingly important to understand when they work well on typical data. For example, the performance of hard parameter sharing depends on dataset properties such as their sample sizes. Yet, rigorously formulating such intuition can be challenging.

	This paper studies a fundamental question: How does hard parameter sharing work in high-dimensional linear regression, where the sample size and feature dimension become increasingly large in a fixed ratio? We present tight generalization bounds for two canonical cases: (i) multiple tasks with the same feature covariates; (ii) two tasks with arbitrarily different sample sizes and covariance matrices. We also demonstrate that these estimates for the empirical loss are incredibly accurate in modest dimensions. Finally, our results provide new explainanations to several intriguing empirical phenomena. For example, increasing one task's sample size helps another task initially by reducing variance, but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

Understanding negative transfer requires developing generalization bounds that scale tightly with properties of each task data, such as its sample size.
This presents a technical challenge in the multi-task setting because of the difference among task features, even for two tasks.
For Rademacher complexity or VC-based techniques, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.
Without a tight lower bound for multi-task learning, comparing its performance to single-task learning results in vacuous bounds.
\todo{add more technical motivation (or maybe later)}
From a practical standpoint, developing a better understanding of multi-task learning in terms of properties of task data can provide guidance for downstream applications \cite{RH19}.
%For example, the sample sizes of all tasks are often assumed to be equal \cite{B00,LPTV09,LPVT11}.
%On the other hand, uneven sample sizes (or dominating tasks) have been empirically observed to cause negative transfer \cite{YKGLHF20}.
%The benefit of learning multi-task representations has also been studied for certain half-spaces \cite{} and sparse regression \cite{}.
%When all tasks are sufficiently similar, adding more labeled data improves the generalization performance for predicting a particular task \cite{WZR20}.

%\textbf{Setup and Main Results.}
In this work, we study the bias and variance of multi-task learning in the high-dimensional linear regression setting \cite{HMRT19,BLLT20}.
Our key observation is that three properties of task data, including \textit{task similarity}, \textit{sample ratio}, and \textit{covariate shift}, can affect whether multi-task learning outperforms single-task learning (which we refer to as \textit{positive transfer}).
As an example, we vary each property in Figure \ref{fig_model_shift_phasetrans} for two linear regression tasks and measure the improvement of multi-task learning over single-task learning for a particular task.
We observe that the effect of transfer can be either positive or negative as we vary each property.
These phenomena cannot be explained using previous techniques \cite{WZR20}.
The high-dimensional linear regression setting allows us to measure the three properties precisely.
Here we define each property for the case of two tasks, while our definition applies to general settings.
We refer to the first task as the source task and the second as the target task.
\squishlist
	\item \textbf{Task similarity:} Assume that both tasks follow a linear model with parameters $\beta_1, \beta_2\in\real^p$, respectively.
	We measure the distance between them by $\norm{\beta_1 - \beta_2}$.
	\item \textbf{Sample ratio:} Let $n_1 = \rho_1 \cdot p, n_2 = \rho_2 \cdot p$ be the sample size of each task, where $\rho_1, \rho_2>1$ are both fixed values that do not grow with $p$.
	We measure the source/target sample ratio by $\rho_1 / \rho_2$.
%	Importantly, $\rho_2$ can be a small constant (say $2$) to capture the need for more labeled data.
	\item \textbf{Covariate shift:} Assume that the task features are random vectors with positive semidefinite covariance matrices $\Sigma_1\in\real^{p\times p}$ and $\Sigma_2\in\real^{p\times p}$, respectively.
	%$x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and is a positive semidefinite matrix.
	We measure covariate shift with matrix $\Sigma_1^{1/2}\Sigma_2^{-1/2}$.
\squishend

%The observations highlight the need to develop generalization bounds that scale tightly with properties of multiple tasks data.
% that only contains limited amount of labeled data.
%Following Hastie et al.  and Bartlett et al. \cite{},

%Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
%The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We consider a multi-task estimator obtained using a shared linear layer for all tasks and a separate output layer for each task \cite{WZR20}.
This two-layer model is inspired by a commonly used idea of hard parameter sharing in multi-task learning \cite{R17,MTDNN19}.
We consider the bias and variance of the multi-task estimator for predicting a target task and compare its performance to single-task learning.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
%Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_prelim}.
%We revisit the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$.

\HZ{go to intro}
For a multivariate Gaussian random matrix, this result follows from the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}.
For a non-Gaussian random matrix, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \citet{bai2009spectral}).
\fi

>>>>

\begin{abstract}
	Hard parameter sharing is a widely used approach to learn jointly from multiple tasks in many applications, such as text classification. The idea is to use a shared feature space for all tasks, while each task also has a separate layer for making prediction.
%	It becomes increasingly important to understand when they work well on typical data. For example, the performance of hard parameter sharing depends on dataset properties such as their sample sizes. Yet, rigorously formulating such intuition can be challenging.
	This paper studies a fundamental question to better understand this approach: How does hard parameter sharing work given multiple linear regression tasks?
	We consider the high-dimensional setting, where the sample size and feature dimension become increasingly large in a fixed ratio, and present tight generalization bounds for two canonical cases: (i) multiple tasks with the same feature covariates; (ii) two tasks with arbitrarily different sample size ratios and covariance matrices. We also demonstrate that these estimates for the empirical loss are incredibly accurate in modest dimensions. Finally, our results provide new explainanations to several intriguing empirical phenomena. For example, increasing one task's sample size helps another task initially by reducing variance, but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

Understanding negative transfer requires developing generalization bounds that scale tightly with properties of each task data, such as its sample size.
This presents a technical challenge in the multi-task setting because of the difference among task features, even for two tasks.
ithout a tight lower bound for multi-task learning, comparing its performance to single-task learning results in vacuous bounds.
\todo{add more technical motivation (or maybe later)}
From a practical standpoint, developing a better understanding of multi-task learning in terms of properties of task data can provide guidance for downstream applications \cite{RH19}.
%For example,
%On the other hand, uneven sample sizes (or dominating tasks) have been empirically observed to cause negative transfer \cite{YKGLHF20}.
%The benefit of learning multi-task representations has also been studied for certain half-spaces \cite{} and sparse regression \cite{}.
%When all tasks are sufficiently similar, adding more labeled data improves the generalization performance for predicting a particular task \cite{WZR20}.

%\textbf{Setup and Main Results.}
In this work, we study the bias and variance of multi-task learning in the high-dimensional linear regression setting \cite{HMRT19,BLLT20}.
Our key observation is that three properties of task data, including \textit{task similarity}, \textit{sample ratio}, and \textit{covariate shift}, can affect whether multi-task learning outperforms single-task learning (which we refer to as \textit{positive transfer}).
As an example, we vary each property in Figure \ref{fig_model_shift_phasetrans} for two linear regression tasks and measure the improvement of multi-task learning over single-task learning for a particular task.
We observe that the effect of transfer can be either positive or negative as we vary each property.
These phenomena cannot be explained using previous techniques \cite{WZR20}.
The high-dimensional linear regression setting allows us to measure the three properties precisely.
Here we define each property for the case of two tasks, while our definition applies to general settings.
We refer to the first task as the source task and the second as the target task.
\squishlist
	\item \textbf{Task similarity:} Assume that both tasks follow a linear model with parameters $\beta_1, \beta_2\in\real^p$, respectively.
	We measure the distance between them by $\norm{\beta_1 - \beta_2}$.
	\item \textbf{Sample ratio:} Let $n_1 = \rho_1 \cdot p, n_2 = \rho_2 \cdot p$ be the sample size of each task, where $\rho_1, \rho_2>1$ are both fixed values that do not grow with $p$.
	We measure the source/target sample ratio by $\rho_1 / \rho_2$.
%	Importantly, $\rho_2$ can be a small constant (say $2$) to capture the need for more labeled data.
	\item \textbf{Covariate shift:} Assume that the task features are random vectors with positive semidefinite covariance matrices $\Sigma_1\in\real^{p\times p}$ and $\Sigma_2\in\real^{p\times p}$, respectively.
	%$x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and is a positive semidefinite matrix.
	We measure covariate shift with matrix $\Sigma_1^{1/2}\Sigma_2^{-1/2}$.
\squishend
\fi

Hard parameter sharing (HPS) is a widely used approach to learn from multiple tasks and goes back to the seminal work of \citet{C97}.
It is generally applied by sharing the feature layers between all tasks, while keeping an output layer for every task.
Recent work has revived interest in this classical approach through many applications \cite{MTDNN19,ZSSGM18}.
Hard parameter sharing offers two critical advantages:
(i) it reduces model parameters since all tasks use the same feature space;
(ii) it reduces the need for labeled data from each task while augmenting the entire training dataset.

Hard parameter sharing is often thought of as a mechanism for inducing inductive bias.
Since all tasks share the same feature space, by restricting the size of the space, HPS acts as a regularizer and reduces overfitting \cite{KD12,WZR20}.
Another source of inductive bias comes from the tasks, and depends on datasets' properties such as sample sizes and task covariances \cite{WZR20}.
Previous generalization theories have considered when all tasks' sample sizes are equal \cite{B00,MPR16}.
Using Rademacher complexity or VC-based techniques, the generalization error scales down as the sample sizes of all tasks increase \cite{AZ05,M06}.
%For, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.

In this paper, we present a new generalization theory for hard parameter sharing by studying its  performance in the high-dimensional linear regression setting.
Linear regression is arguably one of the most fundamental problem in statistics and machine learning.
We are interested in the \textit{high-dimensional} setting, where each dataset's sample size and the feature dimension grow linearly at a fixed ratio.
This is motivated by many multi-task learning applications, where the amount of labeled data from each dataset is usually insufficient for learning a single task.
For example, this is the case if a dataset's sample size is only a small constant factor of the feature dimension \cite{socher2013recursive}.
The high-dimensional setting is challenging but is crucial for understanding how datasets' sample sizes impact generalization performance.

### 2020-10-13 arxiv/setup.tex
polish section 2/3
finish intro

\subsection{Problem Formulation and Main Results}

@@

Regression analysis is arguably one of the most fundamental methods in statistics and machine learning.
In many multi-task learning applications, the amount of labeled data from each dataset is usually insufficient for learning a single task.
For example, this is the case if a dataset's sample size is only a small constant factor of the feature dimension \cite{GLUE}.
In this paper, we consider multi-task learning in a \textit{high-dimensional} linear regression setting, where each dataset's sample size and the feature dimension grow linearly at a fixed ratio.
The high-dimensional setting is challenging but is crucial for understanding how datasets' sample sizes impact generalization performance.
We assume that there are multiple datasets that all follow a (potentially different) linear model.
In each dataset, suppose the feature vector $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with mean zero, unit variance, and covariance $\Sigma \in\real^{p \times p}$ that is deterministic and positive semidefinite.

We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
Given a solution $\hat{A}, \hat{B}$ from the above optimization, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the hard parameter sharing (HPS) estimator for task $i$.
The critical questions here are:
(i) how well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) for datasets with different sample sizes and covariate shifts, how do they impact the estimator?

@@

\paragraph{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (e.g. \cite{hsu2009multi}).
We analyze the global minimum of $f(A, B)$, and provide a sharp generalization bound for its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing variance compared to single-task learning, but hurts by increasing bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample sizes and population covariance matrices.
We analyze the local minimum of $f(A, B)$ and provide a sharp generalization bound for both tasks' prediction loss.
Despite its simplicity, we show several rich phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

@@

\paragraph{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For a matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_t(X)$ denote the eigenvalues of $X$.
Let $X^+$ denote its Moore-Penrose psuedoinverse.
We shall refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices following the standard notations in high-dimensional statistics.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.

>>>>

\subsection{Setup and Main Results}

@@

We assume that there are multiple datasets that all follow a (potentially different) linear model.
In each dataset, suppose the feature vector $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with mean zero, unit variance, and covariance $\Sigma \in\real^{p \times p}$ that is deterministic and positive semidefinite.
We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
Given a solution $\hat{A}, \hat{B}$ from the above optimization, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the hard parameter sharing (HPS) estimator for task $i$.
The critical questions here are:
(i) how well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) for datasets with different sample sizes and covariate shifts, how do they impact the estimator?

@@

\paragraph{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (e.g. \cite{hsu2009multi}).
We analyze the global minimum of $f(A, B)$, and provide a sharp generalization bound for its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing variance compared to single-task learning, but hurts by increasing bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimum of $f(A, B)$ and provide a sharp generalization bound for both tasks' prediction loss.
Despite its simplicity, we show several rich phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

@@

\paragraph{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For a matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_t(X)$ denote the eigenvalues of $X$.
Let $X^+$ denote its Moore-Penrose psuedoinverse.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.
